The illusory promise of the Aligned Rank Transform

### 2025-12-19 appendix_II.qmd
finished methodology

## Absence of spurious effects when $a_2=0$ and $a_{12} = 0${#effects-proof}
Let $x_1$ and $x_2$ be two independent variables, and suppose the conditional distribution of the response $Y | (x_1, x_2)$ depends on these variables only through the linear predictor:
$$
\eta(x_1, x_2) = a_0 + a_1 x_1 + a_2 x_2 + a_{12}x_1 x_2
$$

Assume that the conditional mean can be written as:
$$
\mu(x_1, x_2) = E(Y|x_1, x_2) = h(\eta(x_1, x_2))
$$
where $h$ is a monotone function. This function may coincide with the inverse link function $g^{-1}$ in a GLM, but it may also arise from other frameworks, including transformation models or latent-variable constructions. Thus it does not necessarily need to be smooth or invertible; for example, this assumption is not required for the ordinal example discussed later.

@@

$$\eta(x_1, x_2) = a_0 + a_1x_1$$

@@

$$\frac{\partial \mu(x_1, x_2)}{\partial x_2}  = h'(\eta) \frac{\partial \eta}{\partial x_2} = 0$$
where $h'$ denotes the first derivative of $h$.

@@

$$\frac{\partial^2 \mu(x_1, x_2)}{\partial x_2 \partial x_1} = h''(\eta) \frac{\partial \eta}{\partial x_2} \frac{\partial \eta}{\partial x_1} + h'(\eta)\frac{\partial^2\eta}{\partial x_2 \partial x_1} = 0$$

@@

### Illustrative examples
We discuss two examples.

#### Lognormal data
Let $x_1$ and $x_2$ be two experimental factors. Suppose the true model of the data is
$$
\begin{aligned}
log Y = a_0 + a_1 x_1 + \varepsilon, \qquad
\varepsilon \sim \mathcal{N}(0, \sigma^2)
\end{aligned} 
$$
so that 
$$Y | (x_1, x_2) \sim Lognormal(a_0 + a_1 x_1, \sigma^2)$$

@@

$$Y^* = a_0 + a_1 x_1 + \varepsilon$$

and the observed response $Y$ is obtained by applying fixed thresholds: 
$$
\begin{aligned}
Y = k\ \ \text{if}\ \ \tau_{k-1} < Y^* \leq \tau_k, 
\qquad k = 1,2 ... K
\end{aligned} 
$$ 
Again, the conditional distribution $Y| (x_1, x_2)$ depends on the independent variables only through $\eta(x_1, x_2)=a_0+a_1 x_1$, and there is no effect of $x_2$ on either the latent or the observed ordinal scale.

In this case, the mapping from $\eta$ to the observed response is monotone but not strictly monotone and is therefore non-invertible. As a result, a true effect of $x_2$ on the latent variable (i.e., $a_2 \neq 0$) may be attenuated or entirely unobservable on the ordinal response scale. This does not affect Type I error rates, provided the null hypothesis is defined on the latent scale and the thresholds are fixed, because the absence of a latent effect implies the absence of an effect on the observed scale. However, power and effect sizes can be reduced when effects are assessed on the ordinal scale, reflecting the loss of information due to discretization.

### Invariance under monotone transformations 
Suppose we apply a monotone transformation $T$ directly to the observed responses---such as the rank transformation (RNK) or the inverse normal transformation (INT). Because $T$ is monotone, it preserves the ordering of observations within each condition. The transformed conditional distribution $T(Y) | (x_1, x_2)$ therefore depends on the independent variables only through $\eta(x_1, x_2) = a_0+a_1 x_1$. Consequently, the mean, median, and all quantiles of $T(Y) | (x_1, x_2)$ remain invariant to $x_2$, and marginal effects of $x_1$ do not depend on $x_2$.

>>>>

## Absence of spurious effects when $a_2=0$ and $a_{12} = 0${#effects-proof}
Let $x_1$ and $x_2$ be two independent variables, and suppose the conditional distribution of the response $Y | (x_1, x_2)$ depends on these variables only through the linear predictor:
$$
\ell(x_1, x_2) = a_0 + a_1 x_1 + a_2 x_2 + a_{12}x_1 x_2
$$

Assume that the conditional mean can be written as:
$$
\mu(x_1, x_2) = E(Y|x_1, x_2) = h(\ell(x_1, x_2))
$$
where $h$ is a monotone function. This function may coincide with the inverse link function $g^{-1}$ in a GLM, but it may also arise from other frameworks, including transformation models or latent-variable constructions. Thus it does not necessarily need to be smooth or invertible; for example, this assumption is not required for the ordinal example discussed later.

@@

$$\ell(x_1, x_2) = a_0 + a_1x_1$$

@@

$$\frac{\partial \mu(x_1, x_2)}{\partial x_2}  = h'(\ell) \frac{\partial \ell}{\partial x_2} = 0$$
where $h'$ denotes the first derivative of $h$.

@@

$$\frac{\partial^2 \mu(x_1, x_2)}{\partial x_2 \partial x_1} = h''(\ell) \frac{\partial \ell}{\partial x_2} \frac{\partial \ell}{\partial x_1} + h'(\ell)\frac{\partial^2\ell}{\partial x_2 \partial x_1} = 0$$

@@

### Examples
We discuss two examples.

#### Log-normal distributions
Let $x_1$ and $x_2$ be two experimental factors. Suppose the true model of the data is
$$
\begin{aligned}
log Y = a_0 + a_1 x_1 + \varepsilon, \qquad
\varepsilon \sim \mathcal{N}(0, \sigma^2)
\end{aligned} 
$$
so that 
$$Y | (x_1, x_2) \sim Lognormal(a_0 + a_1 x_1, \sigma^2)$$

@@

$$Y^* = a_0 + a_1 x_1 + \varepsilon, \qquad
\varepsilon \sim \mathcal{N}(0, \sigma^2)$$

and the observed response $Y$ is obtained by applying fixed thresholds: 
$$
\begin{aligned}
Y = d\ \ \text{if}\ \ \tau_{d-1} < Y^* \leq \tau_k, 
\qquad d = 1,2 ... D
\end{aligned} 
$$ 
Again, the conditional distribution $Y| (x_1, x_2)$ depends on the independent variables only through $\ell(x_1, x_2)=a_0+a_1 x_1$, and there is no effect of $x_2$ on either the latent or the observed ordinal scale.

In this case, the mapping from $\ell$ to the observed response is monotone but not strictly monotone and is therefore non-invertible. As a result, a true effect of $x_2$ on the latent variable (i.e., $a_2 \neq 0$) may be attenuated or entirely unobservable on the ordinal response scale. This does not affect Type I error rates, provided the null hypothesis is defined on the latent scale and the thresholds are fixed, because the absence of a latent effect implies the absence of an effect on the observed scale. However, power and effect sizes can be reduced when effects are assessed on the ordinal scale, reflecting the loss of information due to discretization.

### Invariance under monotone transformations 
Suppose we apply a monotone transformation $T$ directly to the observed responses---such as the rank transformation (RNK) or the inverse normal transformation (INT). Because $T$ is monotone, it preserves the ordering of observations within each condition. The transformed conditional distribution $T(Y) | (x_1, x_2)$ therefore depends on the independent variables only through $\ell(x_1, x_2) = a_0+a_1 x_1$. Consequently, the mean, median, and all quantiles of $T(Y) | (x_1, x_2)$ remain invariant to $x_2$, and marginal effects of $x_1$ do not depend on $x_2$.

### 2025-12-19 sandbox.qmd
finished methodology

1. *Normal distribution* $\mathcal{N}(\mu, \sigma)$, where $\mu$ is its mean and $\sigma^2$ is its standard deviation. For most experiments where variances are equal, we set $\sigma = 1$.
2. *Log-normal distribution* $\mathcal{LogN}(\mu, \sigma)$, where $\mu$ is its mean and $\sigma^2$ is its standard deviation at the logarithmic scale. The log-normal distribution is a good model for various measures bounded by zero, such as task-completion times. We set $\sigma = 1$ in our main experiments but evaluate a wider range of parameters in [Appendix I](appendix_I.html#lognormal).
3. *Exponential distribution* $Exp(\lambda)$, where $\lambda$ is its rate. The exponential distribution naturally emerges when describing the time elapsed between events. For example, we could use it to model the time a random person spends with a public display, or the waiting time before a new person approaches to interact with the display, when the average waiting time is $\frac{1}{\lambda}$.  
4. *Cauchy distribution* $\mathcal{Cauchy}(x_0, \gamma)$, where $x_0$ defines its location (median) and $\gamma$ defines its scale. The Cauchy distribution is the distribution of the ratio of two independent normally distributed random variables. It rarely emerges in practice. However, it is commonly used in statistics to test the robustness of statistical procedures because both its mean and variance are undefined. As we discussed earlier, past evaluations of ART [@Mansouri:1995; @elkin:2021] show that the method fails under the Cauchy distribution. We set $\gamma = 1$.
5. *Poisson distribution* $\mathcal{Pois}(\lambda)$, where $\lambda$ is its rate. It expresses the probability of a given number of events in a fixed interval of time. For example, we could use it to model the number of people who interact with a public display in an hour, when the average rate is $\lambda$ people per hour.
6. *Binomial distribution* $\mathcal{B}(\kappa, p)$, where $\kappa$ defines the number of independent Bernoulli trials and $p$ is the probability of error or success of each trial. It frequently appears in HCI research, as it can model the number of successes and failures in a series of experimental tasks. We set $\kappa = 10$ in our main experiments but evaluate a wider range of parameters in [Appendix I](appendix_I.html#binomial).
7. Distributions of *Likert-item* responses with 5, 7, and 11 levels.

**Experimental designs.** We present results for five experimental designs. To simplify our presentation, we start with a 4 $\times$ 3 within-subjects factorial design. We then show how our conclusions generalize to four additional designs: (i) a 2 $\times$ 3 between-subjects design; (ii) a 2 $\times$ 4 mixed design, with a between-subjects factor and a within-subjects factor; (iii) a 2 $\times$ 2 $\times$ 2 within-subjects design, and (iv) a 3 $\times$ 3 $\times$ 3 within-subjects design.

@@

**Equal vs. unequal variances.** We test the robustness of the methods when the variances on the latent variable are unequal.

**Evaluation measures.** In addition to Type I error rates, we compare the statistical power of the methods and assess the accuracy of their effect size estimates.

@@

$$ 
\eta_{ijk} = \mu + s_k + a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j}
$${#eq-linear-model}

@@

**Generating ratio data.** We adopt generalized linear modeling for ratio data, where we use an inverse link function $g^{-1}$ to make the connection between $\eta_{ijk}$ and the location parameter of the distribution of interest:  
$$
Y_{ijk} \sim \mathcal{F}(location=g^{-1}(\eta_{ijk}), ...)
$${#eq-glm-ratio}

where $\mathcal{F}$ denotes the probability density (or mass) function from which we draw samples. For example, the mean of the exponential distribution is equal to $1 / \lambda$, where $\lambda$ is its rate parameter. This mean is linked to the linear term through the function $g^{-1}(x) = e^{x}$, and therefore, $\lambda = e^{-\eta_{ijk}}$.

 @tbl-distributions presents how the parameters of the other distributions are linked to the linear term.

| distribution | parameters | description               |
|--------------|------------|---------------------------|
| Normal       |$\mu = \eta_{ijk}$ <br> $\sigma^2 = \sigma_{\varepsilon}^2$ | mean <br> standard deviation |
| Log-normal    |$\mu = \eta_{ijk}$ <br> $\sigma^2 = \sigma_{\varepsilon}^2$ | mean at logarithmic scale (logarithm of median) <br> standard deviation at logarithmic scale|
| Exponential  |$\lambda = e^{-\eta_{ijk}}$ | rate (1 / mean) |
| Cauchy       |$x_0 = \eta_{ijk}$ <br> $\gamma = 1$ | location (median) <br> scale 
| Poisson      |$\lambda = e^{\eta_{ijk}}$ | rate (mean) |
| Binomial     |$p = Logistic(\eta_{ijk})$ <br> $\kappa = 10$ | probability of error <br> number of Bernoulli trials |

: Distribution parameters and their links to the linear term $\eta_{ijk}$ {#tbl-distributions}

**Generating ordinal data.**

### Population control and distribution conversions 
To simplify our simulations, we fix the following population parameters: $\mu = 0$, $\sigma = 1$, and $\sigma_s = 1$. We then control the magnitude of effects by varying $a_1$, $a_2$, and, for some experiments $a_{12}$. @fig-effects presents the range of values that we test for $a_1$ and $a_2$. We also visualize their effects on the latent variable for factors with two, three, and four categorical levels.

::: {#fig-effects}
```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 9, warning=FALSE}
source("effects_plot.R")
plotEffects()
```
Varying $a_1$ or $a_2$ to control the magnitude of main effects. The plots show examples of population distributions on the latent variable $\mathcal{Y}$ for factors ($X_1$ or $X_2$) with two (top), three (middle), or four categorical levels (bottom).    
:::

We follow the approach of @DeBruine:2021 and use the R package *faux* v1.2.1 [@faux] to simulate data for our mixed-effects models. We assume that the distributions of random intercepts and errors are normal, or $s_k \sim N(0,\sigma_s)$ and $\varepsilon_{ijk} \sim N(0,\sigma)$.
To simulate the observed variable $Y$, we then transform the response values $y_{ijk}$ as described in [Section 3](#approach). A key advantage of this method is that we can generate responses for any distribution, while we control effects on the latent variable in the exact same way.

### Implementation of rank transformation methods
For the aligned rank transformation (ART), we use the R implementation of ARTool v0.11.1 [@artool]. For the pure rank transformation (RNK), we use R's *rank()* function. We use the *Rankit* formulation [@Bliss:1956] for the inverse normal transformation (INT), as explained earlier. Unless explicitly mentioned otherwise, we use the formula ``Y' ~ X1*X2 + (1|S)`` with R's *lmer* function, except for the between-subjects design where we use the formula ``Y' ~ X1*X2 + Error(S)`` with R's *aov* function.

>>>>

1. *Normal distribution* $\mathcal{N}(\mu, \sigma^2)$, where $\mu$ is its mean and $\sigma^2$ is its standard deviation. For most experiments where variances are equal, we set $\sigma^2 = 1$.
2. *Log-normal distribution* $\mathcal{LogN}(\mu, \sigma^2)$, where $\mu$ is its mean and $\sigma^2$ is its standard deviation at the logarithmic scale. The log-normal distribution is a good model for various measures bounded by zero, such as task-completion times. We set $\sigma = 1$ in our main experiments but evaluate a wider range of parameters in [Appendix I](appendix_I.html#lognormal).
3. *Exponential distribution* $Exp(\lambda)$, where $\lambda$ is its rate. The exponential distribution naturally emerges when describing the time elapsed between events. For example, we could use it to model the time a random person spends with a public display, or the waiting time before a new person approaches to interact with the display, when the average waiting time is $\frac{1}{\lambda}$.  
4. *Cauchy distribution* $\mathcal{Cauchy}(x_0, \gamma)$, where $x_0$ defines its location (median) and $\gamma$ defines its scale. The Cauchy distribution is the distribution of the ratio of two independent normally distributed random variables. It rarely emerges in practice. However, it is commonly used in statistics to test the robustness of statistical procedures because both its mean and variance are undefined. As we discussed earlier, past evaluations of ART [@Mansouri:1995; @elkin:2021] show that the method fails under the Cauchy distribution. We set $\gamma = 1$.
5. *Poisson distribution* $\mathcal{Pois}(\lambda)$, where $\lambda$ is its rate. It expresses the probability of a given number of events in a fixed interval of time. For example, we could use it to model the number of people who interact with a public display in an hour, when the average rate is $\lambda$ people per hour.
6. *Binomial distribution* $\mathcal{B}(\kappa, p)$, where $\kappa$ defines the number of independent Bernoulli trials and $p$ is the probability of error or success of each trial. It frequently appears in HCI research, as it can model the number of successes and failures in a series of experimental tasks. We set $\kappa = 10$ in our main experiments but evaluate a wider range of parameters in [Appendix I](appendix_I.html#binomial).
7. Distributions of *Likert-item* responses with 5, 7, and 11 discrete levels.

**Experimental designs.** We present results for five experimental designs. To simplify our presentation, we start with (i) a 4 $\times$ 3 within-subjects factorial design. We then show how our conclusions generalize to four additional designs: (ii) a 2 $\times$ 3 between-subjects design; (iii) a 2 $\times$ 4 mixed design, with a between-subjects factor and a within-subjects factor; (iv) a 2 $\times$ 2 $\times$ 2 within-subjects design; and (v) a 3 $\times$ 3 $\times$ 3 within-subjects design.

@@

**Equal vs. unequal variances.** For normal and ordinal distributions, we test the robustness of the methods when variances are unequal.

**Evaluation measures.** In addition to Type I error rates, we compare the statistical power of the methods and compare their effect size estimates with ground-truth estimates.

@@

$$ 
\ell_{ijk} = \mu + s_k + a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j}
$${#eq-linear-model}

@@

**Generating ratio data.** We adopt generalized linear modeling [@Bolker2009glm] for ratio data, where we use an inverse link function $g^{-1}$ to make the connection between $\ell_{ijk}$ and the location parameter of the distribution of interest:  
$$
Y_{ijk} \sim \mathcal{F}(location=g^{-1}(\ell_{ijk}), ...)
$${#eq-glm-ratio}

where $\mathcal{F}$ denotes the probability density (or mass) function from which we draw the responses $Y_{ijk}$. For example, the mean of the exponential distribution is equal to $1 / \lambda$, where $\lambda$ is its rate parameter. This mean is linked to the linear term through the function $g^{-1}(x) = e^{x}$, and therefore, $\lambda = e^{-\ell_{ijk}}$. Or the link function commonly used for the binomial distribution is the *logit* function, but its inverse if the *logistic* function $g^{-1}(x) = 1 / (1 + e^{-x})$.

 @tbl-distributions presents how the location parameters of the other distributions are linked to the linear term. We emphasize that the Cauchy distribution does not define a GLM (since it does not belong to the exponential family), but we use a model that connects its location (median) through the identity link.

| distribution | location parameter | description               |
|--------------|------------|---------------------------|
| Normal       |$\mu = \ell_{ijk}$ | mean |
| Log-normal    |$\mu = \ell_{ijk}$ | mean at logarithmic scale (logarithm of median) |
| Exponential  |$\lambda = e^{-\ell_{ijk}}$ | rate (1 / mean) |
| Cauchy       |$x_0 = \ell_{ijk}$ | median |
| Poisson      |$\lambda = e^{\ell_{ijk}}$ | rate (mean) |
| Binomial     |$p = 1 / (1 + e^{-\ell_{ijk}})$ | probability of success or error |

: Distribution parameters and their links to the linear term $\ell_{ijk}$ {#tbl-distributions}

**Generating ordinal data.** For generating ordinal data, we follow a latent-variable approach as explained by @Liddell:2018. Specifically, we consider that there is a continuous latent variable that follows a normal distribution: 
$$
Y_{ijk}^* \sim \mathcal{N}(\mu = \ell_{ijk}, \sigma^2 = 1)
$${#eq-ordinal-latent}
From this, we derive discrete responses $d = 1,2 ... D$ by applying fixed thresholds $\tau_d \in (-\infty, \infty)$ as follows:
$$
\begin{aligned}
Y_{ijk} = d\ \ \text{if}\ \ \tau_{d-1} < Y_{ijk}^* \leq \tau_d
\end{aligned} 
$${#eq-ordinal-thresholds}
Since, in practice, $\tau_0 = -\infty$ and $\tau_D = \infty$, generating responses with $D$ discrete levels requires $D-1$ thresholds.

### Effects and population control
We next detail how we select the parameters of the linear component (see @eq-linear-model) for the various distributions, as well as the thresholds of the ordinal model (see @eq-ordinal-thresholds).

**Magnitude of effects.** We control the magnitude of effects by varying $a_1$, $a_2$, and, for some experiments (those evaluating power for interactions), $a_{12}$. @fig-effects-normal presents normal distributions ($\sigma^2 = 1$) generated across the range of values of $a_1$ used to evaluate Type I error rates. In these examples, we illustrate effects for a factor with three categorical levels and, for clarity, disregard between-subject variability ($\sigma_k^2 = 0$).

::: {#fig-effects-normal}
```{r, echo=FALSE, message=FALSE, fig.height=1.2, fig.width = 9, warning=FALSE}
source("distributions_plots.R")
plot_effects_normal()
```
Varying $a_1$ to control the magnitude of the main effect of $X_1$. The plots show examples of normal distributions ($\sigma^2=1$) for a factor with three categorical levels. 
:::

**Grand mean and variance parameters.** We select the mean term $\mu$ to match a target location of the responses. @tbl-locations reports the target location for each distribution, which is typically the overall mean (or the median in the case of the Cauchy distribution). We describe how $\mu$ is derived from these parameters in [Appendix II](appendix_II.html#mu).

| Normal | Log-normal | Exponential | Cauchy | Poisson | Binomial | Ordinal |
|:------:|:----------:|:---------:|:-------:|:--------:|:--------:|:--------:|
| 0 | 1 | 0.5 | 0 | 3 | 0.1 (per trial) | 0 |

: Target mean (median for the Cauchy distribution) of responses $Y_{ijk}$ {#tbl-locations}

@fig-effects-distributions shows distributions generated with these location values when $a_1 = 2$ for a factor with three categorical levels.

::: {#fig-effects-distributions}
```{r, echo=FALSE, message=FALSE, fig.height=1.2, fig.width = 9, warning=FALSE}
source("distributions_plots.R")
plot_effects_distr()
```
Different ratio distributions generated with $a_1 = 2$ when $X_1$ has three categorical levels (here, $\sigma_k^2 = 0$).
:::

In practice, we randomly choose the standard deviation $\sigma_k^2$ --- which determines between-subject variability --- from the interval $[0.1, 0.5]$, drawing each value from a uniform distribution.

**Thresholds of ordinal scales.** For ordinal data, we evaluate $D =$ 5, 7, or 11 discrete levels. To derive the discretization thresholds $\tau_d$, we first consider the range $[-2SD, 2SD]$ , where $SD$ is the overall standard deviation of the latent variable $Y^{*}_{ijk}$. We then divide this range into 5, 7, or 11 intervals, following two different strategies: (i) setting thresholds to be *equidistant*; or (ii) considering *flexible* thresholds, randomly drawing their position in the above range.

 @fig-ordinal presents examples of equidistant and flexible thresholds for a 5-level scale when the magnitude of main effect of $X_1$ is either $a_1 = 2$ or $a_1 = 8$, while all other effects are zero.

::: {#fig-ordinal}
```{r, echo=FALSE, message=FALSE, fig.height=2.5, fig.width = 9, warning=FALSE}
plot_thresholds()
```
The four vertical lines in each plot represent thresholds defining 5-level ordinal scales. Thresholds are either equidistant (top) or flexible (bottom) within a range of $\pm 2$ standard deviations around the grand mean $\mu = 0$.  
:::

### Implementation of rank transformation methods
For the aligned rank transformation (ART), we use the R implementation of ARTool v0.11.1 [@artool]. For the pure rank transformation (RNK), we use R's *rank()* function. We use the *Rankit* formulation [@Bliss:1956] for the inverse normal transformation (INT), as explained earlier.

Unless stated otherwise, we analyze the data using ANOVA with R’s *aov()* function and the model specification ``Y' ~ X1 * X2 + Error(S)``. For balanced designs, this approach yields results that are identical or nearly identical to those obtained using R's *lmer()* function with the model specification ``Y' ~ X1 * X2 + (1 | S)``.
